Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-20 papers/clusters.tex
Anna Y. Q. Ho

Part of the reason that chemical tagging remains unrealized
is because the level of abundance specificity required
is very high.
If there are thousands of (relevant) molecular clouds forming stars in
the recent history of the Milky Way, clumps of stars can only be
identified in abundance space if abundance space is high in dimensionality.
(In principle, it needs to be high in dimensionality both in terms
of the number of measured abundances \emph{and} in terms of the number of
nucleosynthetic pathways or the dimensionality of the true abundance
space.)
Therefore accurate---or at
least precise---measurements of many different abundances are needed
for stellar sibling to have sufficiently unique fingerprints.
Stellar spectroscopic surveys now have the resolution,
$S/N$, wavelength coverage, and sample sizes to deliver many
different chemical tags for each star.

There are, however, two big issues.
The first is that the physical assumptions behind the idea may require
refinement:
There may be chemical-abundance overlaps among open clusters
\citep{blancocuaresma}; coeval groups of stars may have similar tags
but different birth places \citep{mitschang}, and the
chemical-abundance space might be low in dimensionality.
On the other hand, precise studies of stellar twins \citep{melendez, jofre}
and open clusters \citep{bovy} indicates that chemical abundances ought to be
very precisely similar for co-eval stars.

@@

Our specific contribution in this space has been to develop
\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
stellar spectra.
This model can deliver stellar parameters and chemical abundances for
stars, making use of every pixel of every stellar spectrum (that is,
all the information in the data) but making no use at all of physical
models of stars.
It relies only on there being training data---\emph{some} stars for
which parameters and abundances are known and believed.
In companion papers \citep{casey16, ness16} we show that
\thecannon\ can deliver 15 to 19 abundances for stars in the
\apogee\ survey, at precisions higher than even the training data on
which the model is trained.
We say more about this below.
We will show here that \thecannon\ improves chemical abundance measurements
to the point that \emph{chemical tagging is now possible.}

@@

It is important to realize that analysis techniques based on \foreign{ab-initio}
physical models are inaccurate, yet they too strive to improve precision
in the face of knowingly inaccurate models.  Incomplete atomic data, 
simplifications of photospheric structure, presumptions about convective
motion, and inconsistencies resulting from assumptions of the local thermal
equilibrium all contribute to produce knowingly inaccurate abundances.
\thecannon\ stands out because it demonstrably improves the precision on
chemical abundances, whereas the accuracy of those labels is limited only
by the training set employed.

In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---explores the thin disk.
However, it has the disadvantage that it's spatial coverage is incomplete
(that is, limited pointings produce a fractured sky map),
which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage.

One final note:
we think of this \documentname\ as performing a proof of concept.
We know that the stellar members of open and globular clusters---stars
that are identified by being close in phase space---contain highly
informative abundance information that identifies them also in
chemical-abundance space.  Does this work the other way around? 
Can chemical tagging identify small subsets of stars, among a vastly
greater background sample, that have a common birth origin? If we 
find stars purely by their clustering in abundance space and subsequently
show that they are part of a still spatially coherent cluster, then
we will have resolved all practical outstanding issues plaguing 
chemical tagging, thereby coming closer to the tantalizing potential 
in truly unravelling the formation of the Milky Way.

@@

\begin{figure}[!p]
\insanefigure{all}
\caption{The full sample of \totalnumber\ stars used in this study.
  The top six panels show six two-dimensional projections of the
  empirical abundance space distribution.
  The clustering algorithm (described in
  \sectionname~\ref{sec:method}) performs clustering in a full
  15-dimensional abundance space of elements referenced to H (not Fe),
  but plots are shown here referenced to Fe for familiarity reasons.
  The bottom three panels show stellar meta-data not used in the
  clustering but used to test whether the abundance-space clusters
  are interesting spatially.\label{fig:all}}
\end{figure}

@@

Briefly, the k-means algorithm is the following:
$\bullet$~Start at some inital guess for the locations of $K$ $D$-space cluster centers ($K$
locations in the $D$-dimensional space).
$\bullet$~Assign each point in the
space (each star in our case) to the closest of the $K$ centers.
$\bullet$~Given this assignment of stars to centers, update each center (each of
the $K$ $D$-space positions) by taking the mean of the locations of
the stars assigned to that center.
$\bullet$~Iterate these steps (assignment of points, followed by taking of means) to convergence.
The output of this algorithm is the converged locations of the $K$ centers and the assignments of
all points to those centers.
This algorithm is fast and performs well in practice in problems of
this nature; also we are not the first to use the k-means algorithm in
abundance space \citep{gratton}.

@@

Importantly, the k-means algorithm assigns \emph{every} star to at
least one cluster.
For this reason, there is no sense in which each ``cluster'' returned
by k-means is in any sense an overdensity in abundance space.
In what follows, we only consider high-density clusters---clusters
that are much more dense than average; these ought to represent true
over-densities in abundance space.

@@

Although we show only three stellar clusters in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, and \ref{fig:Sgr}, many
other clusters are visible among our densest abundance-space k-means
clusters.
These include M15, M92, and M107, among others.
We haven't asked yet whether we detect---as abundance-space
overdensities---all of the stellar clusters we expect to find. (Many
clusters within the \apogee\ data set only contain a few plausible 
members, thereby  complicating any inferences we wish to make about 
detection completeness using simple k-means.) 
This \documentname\ is simply a demonstration of the chemical-tagging
concept; an investigation of whether we can construct complete
catalogs of stellar clusters is beyond the scope.

@@

There are a few points to highlight about the abundance anti-correlations
found in globular clusters. First, the detection of these signatures gives
confirmation to our analysis: The training set was not purposefully selected
to contain members of any globular or open cluster. However amongst the test
set, we recover globular cluster members and their anti-correlations 
in chemical abundances. That is to
say we recover peculiar chemical abundances that are not dominant in our
training set, but are expected from previous studies. Secondly, the very
presence of complex abundance correlations makes the detection of clusters
(by k-means) substantially more difficult. Briefly, the k-means algorithm
is most effective for near-normally distributed clusters in high-dimensional space.
While globular clusters often show distributions of this kind in a few dimensions, they also demonstrate non-linear correlations which would be
sub-optimally selected by k-means. For these reasons the fact that (a)~\thecannon\
reports these abundance correlations and (b)~k-means--as a simple algorithm without information about expectations in abundance space--does not appear to be
severely impacted by these correlations, provides high promise for more sophisticated approaches.

@@

\acknowledgements
It is a pleasure to thank
  Jeremy Magland (\acronym{SCDA})
for valuable discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

>>>>

Part of the reason that chemical tagging remains unrealized
is because the level of abundance specificity required
is very high.
If there are thousands of (relevant) molecular clouds forming stars in
the recent history of the Milky Way, clumps of stars can only be
identified in abundance space if abundance space is high in dimensionality.
(In principle, it needs to be high in dimensionality both in terms
of the number of measured abundances \emph{and} in terms of the number of
nucleosynthetic pathways or the dimensionality of the true abundance
space.)
Therefore accurate---or at
least precise---measurements of many different abundances are needed
for stellar siblings to have sufficiently unique fingerprints.
Stellar spectroscopic surveys now have the resolution,
$S/N$, wavelength coverage, and sample sizes to deliver many
different chemical tags for each star.

There are, however, two big issues.
The first is that the physical assumptions behind the idea may require
refinement:
There may be chemical-abundance overlaps among open clusters
\citep{blancocuaresma}, coeval groups of stars may have similar tags
but different birth places \citep{mitschang}, and the
chemical-abundance space might be low in dimensionality.
On the other hand, precise studies of stellar twins \citep{melendez, jofre}
and open clusters \citep{bovy} indicate that chemical abundances ought to be
very similar for co-eval stars.

@@

Our specific contribution in this space has been to develop
\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
stellar spectra.
This model can deliver stellar parameters and chemical abundances for
stars, making use of every pixel of every stellar spectrum (that is,
all the information in the data) but making no use at all of physical
models of stars.
It relies only on there being training data---\emph{some} reference stars for
which parameters and abundances are known and believed.
In companion papers \citep{casey16, ness16} we show that
\thecannon\ can deliver 15 to 19 abundances for stars in the
\apogee\ survey, at precisions higher than even the training (reference) data on
which the model is trained.
We say more about this below.
We will show here that \thecannon\ improves chemical abundance measurements
to the point that \emph{chemical tagging is now possible.}

@@

It is important to realize that analysis techniques based on \foreign{ab-initio}
physical models are inaccurate, yet they too strive to improve precision
in the face of knowingly inaccurate models.  Incomplete atomic data, 
simplifications of photospheric structure, presumptions about convective
motion, and inconsistencies resulting from assumptions of the local thermal
equilibrium all contribute to produce knowingly inaccurate abundances.
\thecannon\ stands out because it demonstrably improves the precision on
chemical abundances, whereas the accuracy of those labels is limited only
by the training (reference) set employed.

In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---explores the thin disk.
However, it has the disadvantage that its spatial coverage is incomplete
(that is, limited pointings produce a fractured sky map),
which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage.

One final note:
we think of this \documentname\ as performing a proof of concept.
We know that the stellar members of open and globular clusters---stars
that are identified by being close in phase space---contain highly
informative abundance information that identifies them also in
chemical-abundance space.  Does this work the other way around? 
Can chemical tagging identify small subsets of stars, among a vastly
greater background sample, that have a common birth origin? If we 
find stars purely by their clustering in abundance space and subsequently
show that they are part of a still spatially coherent cluster, then
we will have resolved all practical outstanding issues plaguing 
chemical tagging, thereby bringing us much closer to
unravelling the formation of the Milky Way.

@@

\begin{figure}[!p]
\insanefigure{all}
\caption{The full sample of \totalnumber\ stars used in this study.
  The top six panels show six two-dimensional projections of the
  empirical abundance space distribution.
  The clustering algorithm (described in
  \sectionname~\ref{sec:method}) performs clustering in a full
  15-dimensional abundance space of elements referenced to H (not Fe),
  but plots are shown here referenced to Fe for familiarity reasons.
  The bottom three panels show stellar meta-data not used in the
  clustering described below.\label{fig:all}}
\end{figure}

@@

Briefly, the k-means algorithm is the following:
$\bullet$~Start at some initial guess for the locations of $K$ $D$-space cluster centers ($K$
locations in the $D$-dimensional space).
$\bullet$~Assign each point in the
space (each star in our case) to the closest of the $K$ centers.
$\bullet$~Given this assignment of stars to centers, update each center (each of
the $K$ $D$-space positions) by taking the mean of the locations of
the stars assigned to that center.
$\bullet$~Iterate these steps (assignment of points, followed by taking of means) to convergence.
The output of this algorithm is the converged locations of the $K$ centers and the assignments of
all points to those centers.
This algorithm is fast and performs well in practice in problems of
this nature; also we are not the first to use the k-means algorithm in
abundance space \citep{gratton}.

@@

Importantly, the k-means algorithm assigns \emph{every} star to at
least one cluster.
For this reason, there is no sense in which \emph{every} ``cluster'' returned
by k-means is a distinct overdensity in abundance space.
In what follows, we only consider high-density clusters---clusters
that are much more dense than average; these ought to represent true
over-densities in abundance space.

@@

Although we show only three stellar clusters in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, and \ref{fig:Sgr}, many
other clusters are visible among our densest abundance-space k-means
clusters.
These include M15, M92, and M107, among others.
We haven't asked yet whether we detect---as abundance-space
overdensities---all of the stellar clusters we expect to find. (Many
clusters within the \apogee\ data set only contain a few plausible 
members, thereby  complicating any inferences we wish to make about 
detection completeness using simple k-means.) 
This \documentname\ is simply a demonstration of the chemical-tagging
concept; a full investigation of whether we can construct complete
catalogs of stellar clusters is beyond our present scope.

@@

There are a few points to highlight about the abundance anti-correlations
found in globular clusters. First, the detection of these signatures gives
confirmation to our analysis: The training (reference) set was not purposefully selected
to contain members of any globular or open cluster. However amongst the test
set, we recover globular cluster members and their anti-correlations 
in chemical abundances. That is to
say we recover peculiar chemical abundances that are not dominant in our
training set, but are expected from previous studies. Secondly, the very
presence of complex abundance correlations makes the detection of clusters
(by k-means) substantially more difficult. Briefly, the k-means algorithm
is most effective for near-normally distributed clusters in high-dimensional space.
While globular clusters often show distributions of this kind in a few dimensions, they also demonstrate non-linear correlations which would be
sub-optimally selected by k-means. For these reasons the fact that (a)~\thecannon\
reports these abundance correlations and (b)~k-means--as a simple algorithm without information about expectations in abundance space--does not appear to be
severely impacted by these correlations, provides high promise for more sophisticated approaches.

@@

\acknowledgements
It is a pleasure to thank
  Anna~Y.~Q.~Ho (Caltech),
  Jeremy~Magland (\acronym{SCDA}), and
  the Blanton--Hogg group meeting
for valuable comments and discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.
